---
output: 
  md_document:
    variant: markdown_github
---

# Forecasting Capital Bikeshare usage with ARIMA


***** 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F,warning = F,message = F,cache = T, error=F, fig.width = 9)

# Libraries
  library(tidyverse)
  library(lubridate)
  library(scales)
  library(tidyquant)
  library(quantmod)
  library(forecast)
  library(tseries)
  library(TSstudio)
  library(ggpubr)

# Reading in NOAA.csv with average temperature data for DC
noaa <- read.csv(file="~/Personal Git/DC-Bike-Share-Forecasting/Monthly Forecast/USW00013743.csv")
noaa <- noaa %>% select(STATION,DATE,LATITUDE,LONGITUDE,ELEVATION,NAME,TAVG,TAVG_ATTRIBUTES)

noaa$TAVG <- (((noaa$TAVG/5)*9)+32) 

noaa$Month <- substr(noaa$DATE,6,7) %>% as.numeric()
noaa$Year <- substr(noaa$DATE,1,4) %>% as.numeric()

noaa <- noaa %>% select(Year,Month,TAVG) 

noaa <- noaa %>% 
  filter(Year>=2010) %>% 
  filter(Year<2019)

noaa <- noaa[10:104, ]  

noaa_ts <- ts(noaa$TAVG,
                  start=c(2010,10), end=c(2018,8), 
                  frequency = 12)
  

# Loading cleaned bike data from Oct, 2010 - Aug, 2018 
  # source: (https://s3.amazonaws.com/capitalbikeshare-data/index.html)

load(file="bike_trips.rdata")
monthly_bike_trips <- bike_trips %>% filter(Date>="2010-10-01") %>% filter(Date<"2018-09-01")
  
monthly_bike_trips <- monthly_bike_trips %>%
    mutate(Year=lubridate::year(Date),
           Month=lubridate::month(Date)) %>%
    group_by(Year,Month) %>%
    summarise(Monthly_Trips=sum(n,na.rm=T))

ts_month <- ts(monthly_bike_trips$Monthly_Trips,
                  start=c(2010,10), end=c(2018,8), 
                  frequency = 12)
  

# links 
  # https://towardsdatascience.com/time-series-analysis-with-auto-arima-in-r-2b220b20e8ba
  # https://people.duke.edu/~rnau/411arim.htm
  # https://towardsdatascience.com/stationarity-in-time-series-analysis-90c94f27322

```



Having used DC Capital BikeShare data [previously](https://rsolter.github.io/r/forecasting/Monthly_Bike_Forecast_ETS/) to forecast with exponential smoothing models, I wanted to do the same with ARIMA modeling. ARIMA a complementary method for forecasting univariate timeseries, but it also allows for explanatory variables in the regression. In this case, I’ve added monthly average temperatures as a regressor. The data was [downloaded](https://www.ncei.noaa.gov/access/search/data-search/global-summary-of-the-month) from NOAA and represents measurements from the Regan National Airport weather station.


****

**Stationarity**

A requirement of ARIMA modeling is stationarity of the series which is achieved by having _time invariant_ mean, variance, and co-variance of the series. Any time series with a trend or seasonality is not stationary. 
  
One tool used to achieve stationarity is **differencing.** or computing the difference between consecutive observations. As an example, we can show this using stock data downloaded using the **quantmod** R package. By differencing the data by just one observation, the trend in the stock prices completely disappears:

```{r differencing example, echo=FALSE}
options("getSymbols.warning4.0"=FALSE)
options("getSymbols.yahoo.warning"=FALSE)

quantmod::getSymbols("AAPL", 
                     from = '2017-01-01',
                     to ='2017-06-01', 
                     warnings=FALSE,auto.assign=TRUE)

closingApple <- AAPL$AAPL.Close
DclosingApple <- diff(closingApple,1)
DclosingApple <- DclosingApple[2:length(DclosingApple)] # removing NA


#layout(matrix(c(1,2),2,1),byrow=TRUE)
graphics::par(mfrow=c(2,1))
plot(closingApple, main ="Closing Prices on Apple Stock")
plot(DclosingApple, main ="Differenced Closing Prices on Apple Stock")

```


A second tool for handline 

```{r}

quantmod::getSymbols("NVDA", 
                     from = '2016-10-01',
                     to ='2017-06-01', 
                     warnings=FALSE,auto.assign=TRUE)

closingNVDA <- NVDA$NVDA.Close
transformed_NVDA <- BoxCox(closingNVDA,lambda = "auto")

graphics::par(mfrow=c(2,1))
plot(closingNVDA, main ="Closing Prices on NVDA Stock")
plot(transformed_NVDA, main ="Log Transformed Closing Prices on NVDA Stock")

```



However, was that enough to achieve stationarity? Formally, stationarity can be assessed by using one of many unit tests, one of which is the Kwiatkowski-Phillips-Schmidt-Shin (**KPSS**) test. The null hypothesis for the KPSS test is that the data are stationary, and we look for evidence that the null hypothesis is false. Consequently, small p-values (e.g., less than 0.05), indicate the data is not stationary and suggest differencing is required. Read more [here](https://nwfsc-timeseries.github.io/atsa-labs/sec-boxjenkins-aug-dickey-fuller.html).

In the first test on the raw data, the null hypothesis of stationarity is rejected. Run a second time on differenced data, the null hypothesis is not rejected, indicating that our differencing worked!

```{r adf test}

tseries::kpss.test(closingApple) # p-value 0f 0.01

tseries::kpss.test(DclosingApple) # p-value 0f 0.1

#urca::ur.kpss(closingApple) %>% summary()  
#urca::ur.kpss(DclosingApple) %>% summary() # p-value > 0.1 - Stationary!

```



**ARIMA Theory**

ARIMA is an acronym for Auto Regressive (AR) Integrated (I) Moving Average (MA) and can be implemented in R using the automated function `auto.arima` from the forecast package. ARIMA modeling is best understood by breaking it into its components:

**Auto-regressive** models forecast using a linear combination of past values of the variable. An AR model of order _p_ can be written as:

$$y_{t} = c + \theta y_{t-1} + \theta y_{t-2} + ... + \theta y_{t-p} + \epsilon_{t}$$

Where $\epsilon$ is white noise and the model is denoted **AR(_p_)**, an auto-regressive model of order p. Read more [here](https://otexts.com/fpp2/AR.html)


**Moving Average** models use past values forecast errors as regressors for forecasting:

$$y_{t} = c + \epsilon_{t} + \theta_{1} \epsilon_{t-1} + \theta_{2} \epsilon_{t-2} + ... + \theta_{q} \epsilon_{t-q}$$
Again, in this model, $\epsilon$ is white noise and the model is denoted **MA(_q_)**, a moving-average model of order q. Read more [here](https://otexts.com/fpp2/MA.html) 



**ARIMA Formulation** 

Combining the autoregression and moving average model, we obtain a non-seasonal ARIMA model which is denoted as **ARIMA**(_p,d,q_)**model**, where 
  
  * p - order of the auto-regressive part
  * d - degree of first differencing involved
  * q - order of the moving average part
  
The **seasonal** version of ARIMA builds upon this to include 4 additional seasonal terms:

  * m - number of observations per year (12 for months, 4 for quarters, etc.)
  * P - order of the seasonal auto-regressive part
  * D - degree of first differencing involved for seasonal observations
  * Q - order of the seasonal moving average part



****

#### Application to DC Bike Share


Now we'll apply the ARIMA model to the monthly bike share data while using the temperature data as a regressor. 

From the initial plot of the bike data, we can tell the data has clear seasonality with many fewer riders in the winter months. At the same time, we see an growth in the number of total riders that has appeared to slow in recent years. In the plot for temperature we see a predicatble seasonality whose peaks and valleys remain almost equal in size throughout the entire time frame: 

```{r partition, echo=FALSE}
#ts_month
ts_month_partition <- TSstudio::ts_split(ts_month,sample.out = 12)
bike_train <- ts_month_partition$train
bike_test <- ts_month_partition$test

#noaa_ts
noaa_ts_partition <- TSstudio::ts_split(noaa_ts,sample.out = 12)
noaa_train <- noaa_ts_partition$train
noaa_test <- noaa_ts_partition$test

```


```{r, Train Plots}
p1 <- autoplot(bike_train) +
  xlab("Year") + ylab("Total Rides") + scale_y_continuous(label=comma) +
  #ggtitle("Monthly bike rentals") + 
  theme_minimal() + xlab("") +
  theme(plot.title = element_text(hjust = 0.5))


p2 <- autoplot(noaa_train) +
  xlab("Year") + ylab("Monthly Temperature") +
  #ggtitle("Monthly Temperature (F)") + 
  theme_minimal() + xlab("") +
  theme(plot.title = element_text(hjust = 0.5)) + 
  labs(caption = "Validation data not included")


figure <- ggarrange(p1, p2,
                    labels = c("Bike Rides", "Temperature"),
                    ncol = 1, nrow = 2)
figure

```

To achieve **stationarity** in our time series, we need to remove all trend and seasonality components from the data. A decomposed version of the series can be visualized using the decompose() function. Since the seasonal fluctuations appear to grow over time we feed a multiplicative argument to the (seasonal component) type parameter:

```{r, echo=T}
decompose(bike_train, type="multiplicative") %>% autoplot() + 
  xlab("Year") +
  ggtitle("Classical multiplicative decomposition
    of monthly bike rentals") + theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) + xlab("")

```

Removing the seasonality alone from the time series using seasadj() does not yield a stationary timeseries:

```{r Removing Seasonality, echo=T}

decomp_mult <- decompose(bike_train)
deseasonal_cnt <- seasadj(decomp_mult) 

deseasonal_cnt %>%
  autoplot() + 
  xlab("Year") + scale_y_continuous(label=comma) +
  ggtitle("Monthly bike rentals with annual seasonality Removed") + theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) + xlab("")

tseries::kpss.test(deseasonal_cnt)
```

Examining the ACF and PACF plots indicate that differencing the series by 1 observation could help:

```{r ACF and PACF plots}

acf(deseasonal_cnt)

pacf(deseasonal_cnt)

```


And running the KPSS test confirms that differencing the data does return a stationary series: 

```{r Searching for Stationarity: Differencing, echo=T}
deseasoned_count_d1 = diff(deseasonal_cnt, differences = 1)
tseries::kpss.test(deseasoned_count_d1) 

#deseasoned_count_d2 = diff(deseasonal_cnt, differences = 2)
#tseries::kpss.test(deseasoned_count_d2) 


```


From the plot, it appears that the differenced, deseasoned data has a stationary mean: 

```{r Searching for Stationarity: Differencing Plot}
autoplot(deseasoned_count_d1) +
  xlab("Year") + scale_y_continuous(label=comma) +
  ggtitle("Differenced, de-seasoned Bike rentals time series") + theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

****

Running ACF, PACF plots of the differeced data to see what values for _q_, _p_ would be for an ARIMA model:

The ACF plot shows significant auto-correlations at lags 1,6,10,12 (_q_)
```{r Differenced ACF}
acf(deseasoned_count_d1, main='ACF for Differenced Series')
```


The PACF plot shows significant partial-correlations at 5,6, and 9 (_p_)
```{r Differenced PACF}
pacf(deseasoned_count_d1, main='PACF for Differenced Series')
```




#### Modeling

Using the findings from the ACF, PACF plots above, an ARIMA(2,1,9) model is chosen. Evaluating the diagnostic plots for the (2,1,9) residuals return a seemingly random residual plot and no significant autocorrelations. 

```{r Fit }
fit <- arima(deseasoned_count_d1, order=c(2,1,9))
fit 
```


```{r Fit Evaluation}
tsdisplay(residuals(fit), lag.max=15, main='(2,1,9) Model Residuals')
```


_auto.arima()_
 
Comparing the ARIMA(2,1,9) to the results from auto.arima we can see the automated function does not account for the autocorrelation at q(9). Furthermore, the AIC is slightyl smaller in the ARIMA(2,1,9) model.

```{r Auto-Arima, echo=T}
auto.arima(deseasoned_count_d1)
fit2 <- auto.arima(deseasoned_count_d1,seasonal=FALSE) 
tsdisplay(residuals(fit2), lag.max=45, main='(2,0,2) Model Residuals')
```



#### Forecasting and Back-Testing

To fully evaluate the model's predictive power, we set aside the last 12 months of the _deseasonal_cnt_ object and compare our predictions to what was actually observed.

```{r Partition and Forecast, echo=F, eval=F}

#hold <- deseasonal_cnt[84:95]

train <- deseasonal_cnt[1:83]
train <- ts(train, start=2011, frequency=12)

fit_no_holdout <- arima(ts(train, start=c(2010,10), frequency=12), order=c(2,0,9))

fcast_no_holdout <- forecast(fit_no_holdout, h = 12)

# Plot 1
hold <- window(deseasonal_cnt, start = c(2017, 9), end = c(2018,8))

fcast_no_holdout %>% autoplot() +
  xlab("Year") + ylab("") + scale_y_continuous(label=comma) +
  ggtitle("Prediction ARIMA (2,0,9)") +
  autolayer(hold, series="Actual",size=2) + theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# https://www.datascience.com/blog/introduction-to-forecasting-with-arima-in-r-learn-data-science-tutorials


```




****

Comparing both datasets, we see similar behavior. Both datasets are clearly seasonal while the bike dataset does report a clear growth trend through 2016 before slowing in the following years.

```{r period over period % growth, eval=F}

lagged_bike <- ts_month %>% as.numeric() %>% lag(1)
bike <- ts_month %>% as.numeric()
bike_mo_perc_growth <- bike/lagged_bike

lagged_temp <- noaa_ts %>% as.numeric() %>% lag(1)
temp <- noaa_ts %>% as.numeric()
temp_mo_perc_growth <- temp/lagged_temp

comp_DF <- data.frame("period" = seq.Date(from=as.Date("2010/10/01"),to = as.Date("2018/08/01"),by = "month"),
           temp=temp_mo_perc_growth,
           bike=bike_mo_perc_growth)

comp_DF_long <- comp_DF %>% gather(key,value,2:3)

ggplot(comp_DF_long,aes(x=period,y=value, colour=key)) + 
  geom_point() + geom_line() + 
  theme_minimal() +
  xlab("") + ylab("% Change") + scale_y_continuous(label=percent) +
  ggtitle("Month by Month % Change") + theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(legend.title = element_blank(), legend.position="bottom")
  

```


Will give auto.arima a shot, using default parameters. Returns a MAPE of 11.8%

```{r auto-arima with NOAA data}
# Fitting model
fit_basic1<- auto.arima(bike_train,xreg=noaa_train)

# Model summary, fit statistics
summary(fit_basic1)

# Forecasting forward
forecast_1<-forecast(fit_basic1,xreg = noaa_test)

# plotting forecasted vs actual
autoplot(bike_train,main = 'Forecast For DC Bikeshare Ridership - Temp Regressor',xlab = 'Year - Month',ylab = 'Bike Ridership')+
  autolayer(forecast_1, "Forecasted Ridership")+
  autolayer(bike_test,series = 'Actual Ridership')

```


Residual inspection:

```{r inspecting residuals}
checkresiduals(fit_basic1)

```

Auto.arima without a regressor

```{r}
# Fitting model
fit_noreg<- auto.arima(bike_train)

# Model summary, fit statistics
summary(fit_noreg)

# Forecasting forward
forecast_0<-forecast(fit_noreg,h = 12)

# plotting forecasted vs actual
autoplot(bike_train,main = 'Forecast For DC Bikeshare Ridership - No Regressor',xlab = 'Year - Month',ylab = 'Bike Ridership')+
  autolayer(forecast_0, "Forecasted Ridership")+
  autolayer(bike_test,series = 'Actual Ridership')
```



Residual inspection:

```{r inspecting residuals2}
checkresiduals(fit_noreg)

```


```{r second auto-arima with NOAA data}
# Fitting model
fit_long<- auto.arima(bike_train,xreg=noaa_train,allowdrift = TRUE,allowmean = TRUE)

# Model summary, fit statistics
summary(fit_long)

# Forecasting forward
forecast_2<-forecast(fit_long,xreg = noaa_test)

# plotting forecasted vs actual
autoplot(bike_train,main = 'Forecast For DC Bikeshare Ridership - 2 ',xlab = 'Year - Month',ylab = 'Bike Ridership')+
  autolayer(forecast_1, "Forecasted Ridership")+
  autolayer(bike_test,series = 'Actual Ridership')

summary(forecast_1)

```

